canaper: Categorical analysis of neo- and paleo-endemism in R

### 2022-09-28 ms/ms.Qmd
Update ms

The `cpr_rand_comm()` function generates a single random community. 
The first argument, `comm` is a community data frame. 
The second, `null_model` is the name of one of the predefined randomization algorithms available in `vegan`.
The remainder of the arguments are particular to specific types of randomization algorithm.
`cpr_rand_comm()` is typically not called by the user directly, but is provided to help users select randomization algorithms and settings.

@@

The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with CANAPER, the test data from Phylocom [@Webb2008]:

@@

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # cpr_endem_cols is a color palette in canaper for endemism colors
  scale_fill_manual(values = cpr_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

b <- ggplot(
  acacia_canape, aes(x = pe_alt_obs, y = pe_obs, color = endem_type)) +
  geom_abline(slope = 1, color = "darkgrey") +
  geom_point() +
  scale_color_manual(values = cpr_endem_cols) +
  labs(
    x = "Phylogenetic endemism on comparison tree",
    y = "Phylogenetic endemism on actual tree"
  ) +
  theme_bw() +
  theme(legend.position = "none")

@@

The color palette used in @fig-acacia-cpr has been provided in `canaper` as `cpr_endem_cols`, and was selected to be distinguishable regardless of color vision type [@Okabe2002].
We recommend using this color palette for maximum accessibility.

@@

@@

When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells.
Differences between the two are mostly due to a greater number of significant sites in the `canaper` results (@tbl-cpr-biod).
Again, this does not mean that one or the other is more correct; rather, the differences in significant endemism likely reflect differences in the null models that produced the random communities.
Here, both the `rand_structured` and `curveball` null models preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.

@@

@@

Computations were carried out on an MacBook Pro (2019) with 16 GB RAM and a 1.7 GHz, four core processor.
Approximate compute times were 1.3 hr for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal laptop computer.

@@

@@

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # cpr_endem_cols is a color palette in canaper for endemism colors
  scale_fill_manual(values = cpr_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

b <- ggplot(
  acacia_canape,
  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
) +
  geom_abline(slope = 1, color = "darkgrey") +
  geom_point() +
  scale_color_manual(values = cpr_endem_cols) +
  labs(
    x = "Phylogenetic endemism on comparison tree",
    y = "Phylogenetic endemism on actual tree"
  ) +
  theme_bw() +
  theme(legend.position = "none")

@@

ggsave(plot = fig, file = "ms/figure_1.pdf", width = 100, height = 200, units = "mm")

>>>>

The `cpr_rand_comm()` function generates a single random community. 
The first argument, `comm` is a community dataframe (or matrix). 
The second, `null_model` is the name of one of the predefined randomization algorithms available in `vegan`.
The remainder of the arguments are particular to specific types of randomization algorithm.
`cpr_rand_comm()` is typically not called by the user directly, but is provided to help users select randomization algorithms and settings.

@@

The following code illustrates use of `cpr_rand_comm()` with a set of example data that comes with `canaper`, the test data from Phylocom [@Webb2008]:

@@

## `cpr_iter_sim()`

```{r}
#| label: acacia-stats
#| include: false
data(acacia)

acacia_bin <- acacia$comm
total_cells <- dim(acacia_bin)[1] * dim(acacia_bin)[2] 
acacia_bin[acacia_bin > 0] <- 1
acacia_non_empty <- 1 - (sum(acacia_bin) / total_cells)
acacia_non_empty <- scales::percent(acacia_non_empty, accuracy = 0.1)
```

`cpr_iter_sim()` is not required for the CANAPE workflow, but is rather a diagnostic function for use with randomization algorithms that work by exchanging cells of the community matrix such as `swap` [@Gotelli2003] or `curveball` [@Strona2014].
Each such exchange is known as an "iteration".
It is generally impossible to know *a-priori* how many iterations are needed to completely randomize a given matrix; this number depends on properties of the matrix and the randomization algorithm.
Generally, larger matrices with more skew (overabundance of zeros) will require more iterations (ADD CITATION).

`cpr_iter_sim()` helps the user determine the minimum number of iterations needed by conducting successive exchanges and recording the similarity between the original matrix and the randomized matrix at each iteration.
The similarity values should initially decrease until an approximate minimum is reached; further iterations will only result in noise around this minimum.
A number slightly larger than the smallest number of iterations needed to reach the approximate minimum value can then be used for randomizing the community.
The following code demonstrates usage of `cpr_rand_test()` with the *Acacia* dataset [@Mishler2014] that comes with `canaper`.
The *Acacia* dataset is large (@tbl-datasets) and highly skewed (`r acacia_non_empty` zeros).

```{r}
#| label: cpr-iter-sim-demo
data(acacia)

# Conduct up to 100,000 iterations (swaps),
# recording similarity every 1000 iterations
iter_sim_res <- cpr_iter_sim(
  comm = acacia$comm,
  null_model = "curveball",
  n_iterations = 100000,
  thin = 1000
)

# Inspect the output
iter_sim_res
```

It is useful to plot the output to identify the minimum number of iterations needed.
Here we do so with the `ggplot2` package.

```{r}
#| label: cpr-iter-sim-plot
#| fig.height: 8
#| eval: false

library(ggplot2)

ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")
```

The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations (@fig-iter-sim), and that the original matrix and randomized matrix are about 96.5% similar.

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors see cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

b <- ggplot(
  acacia_canape, aes(x = pe_alt_obs, y = pe_obs, color = endem_type)) +
  geom_abline(slope = 1, color = "darkgrey") +
  geom_point() +
  scale_color_manual(values = mishler_endem_cols) +
  labs(
    x = "Phylogenetic endemism on comparison tree",
    y = "Phylogenetic endemism on actual tree"
  ) +
  theme_bw() +
  theme(legend.position = "none")

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the  `cpr_make_pal()` function.

@@

```{r}
#| label: tally-cells
#| include: false

# Tally number of significant cells in canper vs. Biodiverse
sig_tally <-
  comp_table %>%
  mutate(
    endem_type = str_replace_all(
      endem_type, "not significant", "not_significant") %>%
      as.factor() %>%
      fct_collapse(significant = c("neo", "paleo", "mixed", "super"))
      ) %>%
  rename(canaper = n_canaper, biodiverse = n_biodiverse) %>%
  pivot_longer(names_to = "analysis", values_to = "n_sites", -endem_type) %>%
  group_by(endem_type, analysis) %>%
  summarize(n_sites = sum(n_sites), .groups = "drop") %>%
  split(.$analysis) %>%
  purrr::map(., ~split(., .$endem_type))
```

@@

When we compared endemism type between the `canaper` and Biodiverse CANAPE results for Australian *Acacia*, they agreed in `r p_agree` of grid-cells (@tbl-cpr-biod).
The total number of significant cells was nearly identical between the two (*n* = `r sig_tally$canaper$significant$n_sites` and `r sig_tally$biodiverse$significant$n_sites` sites, `canaper` and Biodiverse, respectively).
The `rand_structured` and `curveball` null models both preserve total abundance per species and richness of each site (community matrix marginal sums); in that sense, they are relatively conservative null models [@Strona2018], which may explain the relative high agreement between results.
Selection of an appropriate null model is beyond the scope of this paper, but must be considered carefully in any community ecological analysis.

@@

# Load times
tar_load(acacia_biod_elapsed_min)

@@

Computations were carried out on an MacBook Pro (2019) with 16 GB RAM and a 1.7 GHz, four core processor.
Approximate compute times were `r acacia_biod_elapsed_min` min for Biodiverse and `r times$acacia_canape_cpr_1$minutes` min for `canaper` (sequential, i.e., non-parallel mode, for both).
However, when parallel computing was enabled for `canaper` with two cores, compute time dropped to `r times$acacia_canape_cpr$minutes` min (further increases in number of cores did not significantly decrease time).
This demonstrates that `canaper` can efficiently compute moderately sized datasets with a personal laptop computer.

@@

**@fig-iter-sim.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
The *Acacia* dataset [@Mishler2014] was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.

@@

```{r}
#| label: fig-iter-sim
#| fig.height: 4
#| fig-cap: "Testing the number of iterations needed to randomize the Australian *Acacia* dataset."
#| echo: false

fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")

ggsave(plot = fig, file = "ms/figure_1.pdf", width = 150, height = 150, units = "mm")

@@

a <- ggplot(acacia_canape, aes(x = long, y = lat, fill = endem_type)) +
  geom_tile() +
  # use same colors as Misher et al. 2014
  # (for CVD-safe colors could use cpr_endem_cols or cpr_endem_cols_2)
  scale_fill_manual(values = mishler_endem_cols) +
  guides(
    fill = guide_legend(title.position = "top", label.position = "bottom")
  ) +
  coord_fixed() +
  theme(legend.position = "bottom", legend.title = element_blank())

b <- ggplot(
  acacia_canape,
  aes(x = pe_alt_obs, y = pe_obs, color = endem_type)
) +
  geom_abline(slope = 1, color = "darkgrey") +
  geom_point() +
  scale_color_manual(values = mishler_endem_cols) +
  labs(
    x = "Phylogenetic endemism on comparison tree",
    y = "Phylogenetic endemism on actual tree"
  ) +
  theme_bw() +
  theme(legend.position = "none")

@@

ggsave(plot = fig, file = "ms/figure_2.pdf", width = 100, height = 200, units = "mm")
